---
title: "Preliminary Analysis: GSE262796"
subtitle: "Acquired cetuximab resistance in colorectal cancer"
author: "Chai Yi Kang & Jamie Tan Wen Qi (Chen Wenqi)"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

# Introduction

This document accompanies our analysis plan for GSE262796, an RNA-seq dataset investigating acquired resistance to cetuximab in two colorectal cancer cell lines (LIM1215 and DiFi). The dataset comprises 18 samples across six experimental groups: for each cell line, one cetuximab-sensitive parental population and two independently derived resistant populations, each with three biological replicates.

Here we demonstrate engagement with the real data by (1) loading the GEO sample sheet and displaying the experimental design, and (2) running FastQC on representative samples from different conditions.

# Metadata table
We retrieve the sample sheet for GSE262796 directly from GEO using the `GEOquery` package. This gives us the experimental design in a form that can be programmatically inspected before any sequencing data is touched.

```{r load-metadata}
library(GEOquery)
library(knitr)
library(dplyr)

# Download the series metadata. GSEMatrix = TRUE retrieves the
# sample-level data. getGPL = FALSE skips the platform annotation,
# which is irrelevant for RNA-seq (GPLs describe microarray probes).
gse <- getGEO("GSE262796", GSEMatrix = TRUE, getGPL = FALSE)

# getGEO returns a list (one entry per platform used in the series).
# GSE262796 uses a single platform, so we take the first element.
eset <- gse[[1]]

# The sample sheet lives in pData() of the ExpressionSet object.
pheno <- pData(eset)

cat("Number of samples:", nrow(pheno), "\n")
cat("Number of metadata columns (raw):", ncol(pheno), "\n")
```

The raw sheet contains many GEO boilerplate columns. We extract only the fields relevant to the experimental design.
```{r tidy-metadata}
library(stringr)
library(readr)

raw_pheno <- pData(eset)

# Clean core metadata from GEO
meta <- raw_pheno |>
  mutate(
    sample_id = geo_accession,
    cell_line = `cell line:ch1`,
    srx       = str_extract(relation.1, "SRX\\d+"),
    condition = str_extract(title, "(?<=-)(S|R1|R2)(?= )"),
    replicate = as.integer(str_extract(title, "(?<=biol rep )\\d"))
  ) |>
  mutate(group = paste(cell_line, condition, sep = "-")) |>
  select(sample_id, cell_line, condition, replicate, group, srx)

# SRA Run Selector Metadata was manually downloaded and saved locally
runinfo <- read_csv("sra_runinfo.csv", show_col_types = FALSE)

# Join SRR Run IDs onto our metadata
# SRA Run Selector's CSV uses 'Experiment' and 'Run' as column names
meta <- meta |>
  left_join(runinfo |> select(Experiment, Run), by = c("srx" = "Experiment")) |>
  rename(sra_run = Run) |>
  arrange(cell_line, condition, replicate)

kable(meta, caption = "Experimental design for GSE262796 with SRA run IDs")
```

# Quality control

We run FastQC on two representative samples to characterise library quality and inform trimming decisions. Following the biological rationale of our pipeline, we chose one cetuximab-sensitive and one cetuximab-resistant LIM1215 sample — the minimum necessary pair for our core contrast, sequenced in the same batch.

| Role | GSM | Condition | SRA Run |
|------|-----|-----------|---------|
| Reference | GSM8179487 | LIM1215-S rep 1 | SRR28499039 |
| Resistant | GSM8179490 | LIM1215-R1 rep 1 | SRR28499036 |

## Downloading the FASTQs

We download the two runs from SRA using `sra-tools`. The `prefetch → fasterq-dump` pattern is NCBI's recommended workflow: `prefetch` retrieves the run in compressed SRA format (resumable, efficient), then `fasterq-dump` converts it to FASTQ. The `--split-files` flag creates separate `_1.fastq` and `_2.fastq` files for paired-end data.

This chunk takes ~15–30 minutes on a typical home connection. We set `eval: false` so it does not re-execute on every render — the files persist in `fastq/` once downloaded.

```{bash download-fastqs}
#| eval: false

set -e

mkdir -p fastq fastqc_out
cd fastq

for SRR in SRR28499039 SRR28499036; do
  if [ ! -f "${SRR}_1.fastq.gz" ] && [ ! -f "${SRR}.fastq.gz" ]; then
    echo "Fetching $SRR..."
    prefetch "$SRR"
    fasterq-dump "$SRR" --split-files --threads 4
    # Compress to save space
    gzip -f "${SRR}"*.fastq
    # Clean up the intermediate .sra directory
    rm -rf "$SRR"
  else
    echo "$SRR already present, skipping."
  fi
done

ls -lh
```

## Running FastQC

Once the FASTQs are on disk, FastQC is fast — a minute or two per file. We run it on whatever FASTQs landed in `fastq/` and write HTML + zip reports to `fastqc_out/`.

```{bash run-fastqc}
#| eval: false

fastqc fastq/*.fastq.gz -o fastqc_out -t 4
ls fastqc_out/
```