---
title: "Preliminary Analysis: GSE262796"
subtitle: "Acquired cetuximab resistance in colorectal cancer"
author: "Chai Yi Kang & Jamie Tan Wen Qi (Chen Wenqi)"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

# Introduction

This document accompanies our analysis plan for GSE262796, an RNA-seq dataset investigating acquired resistance to cetuximab in two colorectal cancer cell lines (LIM1215 and DiFi). The dataset comprises 18 samples across six experimental groups: for each cell line, one cetuximab-sensitive parental population and two independently derived resistant populations, each with three biological replicates.

Here we demonstrate engagement with the real data by (1) loading the GEO sample sheet and displaying the experimental design, and (2) running FastQC on one replicate from each of the six experimental conditions.

# Metadata table
We retrieve the sample sheet for GSE262796 directly from GEO using the `GEOquery` package. This gives us the experimental design in a form that can be programmatically inspected before any sequencing data is touched.

```{r load-metadata}
library(GEOquery)
library(knitr)
library(dplyr)

# Download the series metadata. GSEMatrix = TRUE retrieves the
# sample-level data. getGPL = FALSE skips the platform annotation,
# which is irrelevant for RNA-seq (GPLs describe microarray probes).
gse <- getGEO("GSE262796", GSEMatrix = TRUE, getGPL = FALSE)

# getGEO returns a list (one entry per platform used in the series).
# GSE262796 uses a single platform, so we take the first element.
eset <- gse[[1]]

# The sample sheet lives in pData() of the ExpressionSet object.
pheno <- pData(eset)

cat("Number of samples:", nrow(pheno), "\n")
cat("Number of metadata columns (raw):", ncol(pheno), "\n")
```

The raw sheet contains many GEO boilerplate columns. We extract only the fields relevant to the experimental design.
```{r tidy-metadata}
library(stringr)
library(readr)

raw_pheno <- pData(eset)

# Clean core metadata from GEO
meta <- raw_pheno |>
  mutate(
    sample_id = geo_accession,
    cell_line = `cell line:ch1`,
    srx       = str_extract(relation.1, "SRX\\d+"),
    condition = str_extract(title, "(?<=-)(S|R1|R2)(?= )"),
    replicate = as.integer(str_extract(title, "(?<=biol rep )\\d"))
  ) |>
  mutate(group = paste(cell_line, condition, sep = "-")) |>
  select(sample_id, cell_line, condition, replicate, group, srx)

# SRA Run Selector Metadata was manually downloaded and saved locally
runinfo <- read_csv("sra_runinfo.csv", show_col_types = FALSE)

# Join SRR Run IDs onto our metadata
# SRA Run Selector's CSV uses 'Experiment' and 'Run' as column names
meta <- meta |>
  left_join(runinfo |> select(Experiment, Run), by = c("srx" = "Experiment")) |>
  rename(sra_run = Run) |>
  arrange(cell_line, condition, replicate)

kable(meta, caption = "Experimental design for GSE262796 with SRA run IDs")
```

# Quality control

We run FastQC on six representative samples to characterise library quality and inform trimming decisions — one replicate from each of the six experimental conditions (both cell lines × sensitive/R1/R2). This gives us coverage of every condition in the design while keeping the download and QC runtime manageable. Replicate 1 is used throughout for consistency.

| Role | GSM | Condition | SRA Run |
|------|-----|-----------|---------|
| Sensitive parental | GSM8179487 | LIM1215-S rep 1 | SRR28499039 |
| Resistant line 1 | GSM8179490 | LIM1215-R1 rep 1 | SRR28499036 |
| Resistant line 2 | GSM8179493 | LIM1215-R2 rep 1 | SRR28499033 |
| Sensitive parental | GSM8179478 | DiFi-S rep 1 | SRR28499048 |
| Resistant line 1 | GSM8179481 | DiFi-R1 rep 1 | SRR28499045 |
| Resistant line 2 | GSM8179484 | DiFi-R2 rep 1 | SRR28499042 |

## Downloading the FASTQs

We download the six runs from SRA using `sra-tools`. The `prefetch → fasterq-dump` pattern is NCBI's recommended workflow: `prefetch` retrieves the run in compressed SRA format (resumable, efficient), then `fasterq-dump` converts it to FASTQ. The `--split-files` flag creates separate `_1.fastq` and `_2.fastq` files for paired-end data.

This chunk takes ~1–2 hours on a typical home connection for all six samples. We set `eval: false` so it does not re-execute on every render — the files persist in `fastq/` once downloaded. In practice we ran the standalone script `download_fastq_all_conditions.sh` from the project root; the chunk below mirrors it exactly for reproducibility.

```{bash download-fastqs}
#| eval: false

set -euo pipefail

mkdir -p fastq
cd fastq

# One SRR per condition, using replicate 1 throughout for consistency
SRRS=(
  SRR28499039   # LIM1215-S  rep 1
  SRR28499036   # LIM1215-R1 rep 1
  SRR28499033   # LIM1215-R2 rep 1
  SRR28499048   # DiFi-S     rep 1
  SRR28499045   # DiFi-R1    rep 1
  SRR28499042   # DiFi-R2    rep 1
)

for SRR in "${SRRS[@]}"; do
  if [ -f "${SRR}_1.fastq.gz" ]; then
    echo ">>> $SRR already present, skipping."
    continue
  fi

  echo ">>> Fetching $SRR"
  prefetch "$SRR"
  fasterq-dump "$SRR" --split-files --threads 4
  gzip -f "${SRR}"_1.fastq "${SRR}"_2.fastq
  rm -rf "$SRR"
  echo ">>> Done with $SRR"
done

echo ">>> All samples ready:"
ls -lh
```

## Running FastQC

Once the FASTQs are on disk, FastQC is fast — a minute or two per file, so roughly 15–20 minutes for all twelve files (6 samples × paired-end). We run it on whatever FASTQs landed in `fastq/` and write HTML + zip reports to `fastqc_out/`.

```{bash run-fastqc}
#| eval: false

mkdir -p fastqc_out
fastqc fastq/*.fastq.gz -o fastqc_out -t 4
ls fastqc_out/
```